---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Jon Nichols"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Monday, February 28 at 7:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1

getwd()
library(tidyverse)
library(agricolae)
library(lubridate)

Lake <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", stringsAsFactors = TRUE)

Lake$sampledate <- as.Date(Lake$sampledate , format = "%m/%d/%y")

#2

mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right")
theme_set(mytheme)


```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer: Temperature = a + b1(Depth) + e
H0: b1 = 0 
Ha: b1 != 0 


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4

Lake.subset <- 
  Lake %>% 
  mutate(month = month(sampledate)) %>%
  filter(month == 07) %>%
  select(lakename:daynum, depth:temperature_C) %>%
  na.omit()
  
summary(Lake.subset$temperature_C)
summary(Lake.subset$depth)


#5

tempvsdepth <- 
  ggplot(Lake.subset, aes(x = depth, y = temperature_C)) +
  geom_point() +
  geom_smooth(method = lm)+
  ylab("Temperature in Celsius")+
  xlab("Depth")+
  ylim(0, 35) 
print(tempvsdepth)


```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: As depth increases, temperature decreases. The distribution of points appear to be linearly correlated up to a depth of 10m, at which point the temperature hits a limit of approximately 5C even as depth continues to increase. 


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7

lake.regression <- lm(data = Lake.subset, temperature_C ~ depth)
summary(lake.regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: Depth explains 73.87% of the variability in temperature. There are 9,726 degrees of freedom on which this finding is based. The p-value for depth is less than 0.05, indicating that we can reject the null hypothesis and our coefficient for the relationship between depth and temperature is statistically different than zero. The p-value for the entire regression is also lower than the 0.05 confidence level, indicating the regression is meaningful and can help explain the relationship between temperature and depth. 


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9

NTLAIC <- lm(data = Lake.subset, temperature_C ~ depth + year4 + daynum)
step(NTLAIC)



#10

AICmodel <- lm(data = Lake.subset, temperature_C ~ depth + year4 + daynum)
summary(AICmodel)



```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: Depth, year, and day number were included in the final set of explanatory variables. The model explains 74.12% of the observed variance in temperature. This is only a slight improvement over using depth as the sole explanatory variable, and the increase in the R2 is likely due to adding two additional independent variables rather than accounting for additional variance. 



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12

Lake.Totals <- Lake.subset %>%
  group_by(lakename, year4,daynum)%>%
  summarise(meantemp = mean(temperature_C))

Lake.Totals.anova <- aov(data = Lake.Totals, meantemp ~ lakename)
summary(Lake.Totals.anova)

Lake.Totals.anova2 <- lm(data = Lake.Totals, meantemp ~ lakename)
summary(Lake.Totals.anova2)


```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, both anova tests have p-levels less than 0.05, thus we reject the null hypothesis. 



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.

tempvsdepth2 <- 
  ggplot(Lake.subset, aes(x = depth, y = temperature_C, color=lakename)) +
  geom_point(alpha = .5) +
  geom_smooth(method = lm, se=FALSE)+
  labs(color = "Lake Name")+
  ylab("Temperature in Celsius")+
  xlab("Depth")+
  ylim(0, 35) 
print(tempvsdepth2)



```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15

TukeyHSD(Lake.Totals.anova)
Lake.totals.groups <- HSD.test(Lake.Totals.anova, "lakename", group = TRUE)
Lake.totals.groups


```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Ward Lake and Paul Lake 

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We can do a two tailed t-test. Null hypothesis would be mean(peter) - mean(paul) = 0. Alternative hypothesis would be that the mean is not equal to 0. 


